---
format:
  html:
    theme: cosmo
    toc: true
    toc-location: right
    page-layout: full
    css: styles.css
editor_options: 
  chunk_output_type: console
---

<div style="text-align: left;">

# Mini Project 1 Interactive Maps

#### Author: Tenzin Gyaltsen

## Violent Crime Rate Interactive Map (1999)

Explore how violent crime rates varied across U.S. states in 1999 using this interactive map. Hover over any state to view its crime rate per 100,000 residents.

- 📘 **Source title**: [More Guns, Less Crime?](https://vincentarelbundock.github.io/Rdatasets/datasets.html)  
- 🔎 **About this dataset**: [Guns documentation](https://vincentarelbundock.github.io/Rdatasets/doc/AER/Guns.html)

```{r}
#| echo: true
#| message: false
#| warning: false

# Load necessary libraries
library(tidyverse)
library(sf)
library(leaflet)
library(stringr)
library(readr)
library(htmltools)

# ----------- Violent Crime Rate Map (1999) -----------

# Load and filter Guns dataset for 1999
Guns_1999 <- read_csv("Guns.csv") |> 
  filter(year == 1999) |> 
  mutate(state = str_to_lower(str_squish(state)))  # Standardize state names

# Load US states geometry data for interactive mapping
states_sf <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |> 
  select(name, geometry) |> 
  mutate(name = str_to_lower(name))  # Standardize state names

# Merge violent crime data with state geometries
data_interactive <- states_sf |> 
  left_join(Guns_1999, by = c("name" = "state"))

# Define numeric color scale for violent crime rates
pal_numeric <- colorNumeric(
  palette = "magma", 
  domain = data_interactive$violent, 
  reverse = TRUE,  # Ensures high values are dark, low values are light
  na.color = "transparent"  # Hide missing values instead of coloring them
)

# Ensure state names are capitalized before using them in tooltips
data_interactive <- data_interactive |> 
  mutate(name = str_to_title(name))  # Capitalize state names

# Generate interactive choropleth map for violent crime rates
leaflet(data_interactive) |> 
  addTiles() |> 
  addPolygons(fillColor = ~pal_numeric(violent),
              weight = 1, 
              color = "white",
              fillOpacity = 0.7,
              label = lapply(
                paste0(
                  "<strong>", data_interactive$name, "</strong><br>",  # Capitalized state name
                  "Violent Crime Rate (1999): <strong>", data_interactive$violent, "</strong> per 100,000"), HTML),  # Convert text to HTML for proper formatting
              labelOptions = labelOptions(
                style = list(
                  "font-weight" = "normal", 
                  "padding" = "6px",
                  "background" = "rgba(255,255,255,0.8)",  # Semi-transparent white background
                  "border" = "1px solid black",
                  "border-radius" = "5px"
                ),
                textsize = "13px",
                direction = "auto"
              )) |>
  addLegend(pal = pal_numeric, 
            values = ~violent, 
            opacity = 0.7, 
            title = "Violent Crime Rate (1999)", 
            position = "bottomright")
```

📝 **Insight:** States with higher crime rates (darker purple) tend to cluster in the South and Southwest. Many northern states had comparatively lower rates.

## Seat Belt Law Enforcement Interactive Map (1997)

This interactive map shows seat belt law enforcement types in 1997. You can hover over any state to see if they had primary, secondary, or no enforcement at all.

- 📘 **Source title**: [Effects of Mandatory Seat Belt Laws in the US](https://vincentarelbundock.github.io/Rdatasets/datasets.html)  
- 🔎 **About this dataset**: [USSeatBelts documentation](https://vincentarelbundock.github.io/Rdatasets/doc/AER/USSeatBelts.html)

```{r}
#| echo: true
#| message: false
#| warning: false

# ----------- Seat Belt Law Enforcement Map (1997) -----------

# Load and filter seat belt law data for 1997
USSeatBelts_1997 <- read_csv("USSeatBelts.csv") |> 
  filter(year == 1997) |> 
  mutate(state = state.name[match(state, state.abb)],  # Convert abbreviations to full names
         state = str_to_lower(str_squish(state))) |>  # Standardize state names
  drop_na(state)  # Ensure valid state names only

# Load US states geometry data for interactive mapping
states_sf <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |> 
  select(name, geometry) |> 
  mutate(name = str_to_lower(name))  # Standardize state names

# Merge seat belt law data with state geometries
data_interactive <- states_sf |> 
  left_join(USSeatBelts_1997, by = c("name" = "state")) |> 
  mutate(enforce = replace_na(enforce, "NA"))  # Explicitly label missing enforcement data

# Define categorical color mapping for seat belt laws
pal_categorical <- colorFactor(
  palette = c("gray", "yellow", "blue", "red"),  # Matches static map color scheme
  domain = c("no", "primary", "secondary", "NA")
)

# Ensure state names are capitalized before using them in tooltips
data_interactive <- data_interactive |> 
  mutate(name = str_to_title(name))  # Capitalize state names

# Generate interactive choropleth map for seat belt law enforcement
leaflet(data_interactive) |> 
  addTiles() |> 
  addPolygons(fillColor = ~pal_categorical(enforce),
              weight = 1, 
              color = "white",
              fillOpacity = 0.7,
              label = lapply(
                paste0(
                  "<strong>", data_interactive$name, "</strong><br>",  # Capitalized state name
                  "Seat Belt Enforcement: <strong>", data_interactive$enforce, "</strong>"), HTML),  # Convert text to HTML for proper formatting
              labelOptions = labelOptions(
                style = list(
                  "font-weight" = "normal", 
                  "padding" = "6px",
                  "background" = "rgba(255,255,255,0.8)",  # Semi-transparent white background
                  "border" = "1px solid black",
                  "border-radius" = "5px"
                ),
                textsize = "13px",
                direction = "auto"
              )) |>
  addLegend(pal = pal_categorical, 
            values = ~enforce, 
            opacity = 0.7, 
            title = "Seat Belt Law Enforcement (1997)", 
            position = "bottomright")
```

📝 **Insight:** Secondary enforcement was the dominant seat belt policy in 1997. Only a few states had adopted primary enforcement, and New Hampshire stood out with no enforcement.

</div>
